GVN-Hoist: Hoisting Computations from Branches

### 2016-08-09 hoist.tex
time complexity moved down, triangle and pessimistic removed

The complexity of code hoisting is linear in number of instructions that could
be hoisted in the program, matching the complexity of PRE on SSA form.  The
analysis phase is based on the Global Value Numbering (GVN), the same analysis
used for PRE, followed by the computation of a partition of identical
expressions to be hoisted in a same location to guarantee safety properties and
program performance, and followed by a simple code generation that adds the
identified instruction in the hoisting point and removes all the now redundant
expressions.

\subsection{Contributions of this paper}
\begin{itemize}
\item A systematic approach to writing a compiler optimization pass.
\item gvn hoisting
\item partition algorithm
\item cost model to reduce live-range and reduce spills
\end{itemize}

\begin{itemize}
\item Code hoisting in adjacent blocks in a diamond/triangle like structure.
\item Hoisting equivalent computations in non-sibling BBs.
\item Hoisting across switch blocks.
\item Ranking/Wavefront algorithm to hoist across switch blocks.
\end{itemize}

@@

\section{Code hoisting policies}
The amount of hoisting depends on the whether we collect information about all
the instructions before finding candidates (optimistic), or on demand
(pessimistic). It also depends on the generality of the GVN algorithm, however,
that analysis is beyond the scope of this paper.

\subsection{Hoisting bottom up with optimistic approach}
\label{subsec:optimistic}
In this approach, the goal is to maximise the total number of hoistings in the
entire function.  This algorithm is very useful when optimizing for code-size.
In order to do that, we collect the GVN of all the instructions in the function
and iterate on the list of instructions having identical GVNs. After that we
find the common dominator dominating all the identical computations and perform
legality checks (Section-\ref{subsec:legality}). Many times it is not possible
to hoist all the instructions to one common dominator, due to legality
constraints e.g., intersecting side-effects or, profitability constraints e.g.,
hoisting point too far. In those cases, the algorithmn would partition the list
of computations and partially hoist a set of computations to their common
dominator. For more details see Section-\ref{subsec:partition}.

@@

\subsubsection{Triangle approach}
This is the most basic approach to code-hoisting when the instructions are
hoisted to immediate parent in a diamond like structure. When there are
identical computations in both the branches of a conditional, and all the
legality checks are met, the instruction can be hoisted to the predecessor basic
block. This approach is simple to implement and has minimal impact on the
register pressure.

\subsubsection{Pessimistic approach}
\label{subsec:pessimistic}
To extend the triangle approach to the entire function, the algorithm traverses
the basic blocks in the invers depth-first order. All the leaf nodes are visited
before the non-leaf nodes because instructions are hoisted upwards. We start
from a branch and collect the GVN of all the instructions. Then in the sibling
basic block we find if there are any identical computations. Once we find an
identical computation we try to hoist it to the nearest common dominator which
is the immediate predecessor of both the basic blocks. We traverse the basic
block from top to down so that the once an instruction is hoisted it creates
opportunities for other dependent instructions. After all the sibling basic
blocks have all the identical computations hoisted, we continue the hoisting
process with their parent and its sibling of their parent until we reach the
beginning of the function.

\subsubsection{PRE based hoisting}

>>>>

\subsection{Contributions of this paper}
\begin{itemize}
\item hoisting of identical computations
\item partition algorithm
\item code hoisting can remove fully redundant computations
\item code hoisting can actually improve performance in some cases
\item cost model to reduce live-range and reduce spills
\end{itemize}

@@

\section{Code hoisting policies}
The amount of hoisting depends on the whether we collect information about all
the instructions before finding candidates (optimistic).

In the pessimistic approach, the basic blocks are traversed in the inverse
depth-first order, computing the GVN of instructions as they come by. The GVN of
sibling branches are compared for equality. Once such a candidate is found, it
is hoisted in the common dominator. All the leaf nodes are visited before the
non-leaf nodes (bottom-up) because instructions are hoisted upwards. The
pessimistic algorithm is fast, results in fewer spills but hoists very less
instructions. We have implemented optimistic approach because it is more general
and can be tuned down to closely mimic pessimistic approach by changing a flag.

It also depends on the generality of the GVN algorithm, however, that analysis
is beyond the scope of this paper.

\subsection{Hoisting bottom up with optimistic approach}
\label{subsec:optimistic}
In this approach, the goal is to maximise the total number of hoistings in the
entire function.  This algorithm is very useful when optimizing for code-size.
We collect the GVN of all the instructions in the function and iterate on the
list of instructions having identical GVNs. After that we find the common
dominator dominating all such identical computations and perform legality checks
(Section-\ref{subsec:legality}). Many times it is not possible to hoist all the
instructions to one common dominator, due to legality constraints e.g.,
intersecting side-effects or, profitability constraints e.g., hoisting point too
far. In those cases, this algorithmn would partition the list of identical
computations and partially hoist a set of computations to their common
dominator. The partition algorithm is described as follows:

@@

\subsubsection{Time complexity of algorithm}
The complexity of code hoisting is linear in number of instructions that could
be hoisted in the program, matching the complexity of PRE on SSA form.  The
analysis phase is based on the Global Value Numbering (GVN), the same analysis
used for PRE, followed by the computation of a partition of identical
expressions to be hoisted in a same location to guarantee safety properties and
program performance, and followed by a simple code generation that adds the
identified instruction in the hoisting point and removes all the now redundant
expressions.
